---
title: "2022 Presidential Elections Data"
output: html_notebook
---

This is a personal project that aims to analyze the Philippine 2022 Presidential Election using the programming language, R. This project is based on Project Raven's Tableau [project](https://www.facebook.com/61562294530876/posts/pfbid02bwLWd45JfRSiCbmaFYBJRfM4WbYD6diK8rBAHRzoDjDwoZF8DWW8HZJ6gKJRhp9tl/). 

My personal aim in achieving this project is to avoid using any AI for assistance to fully immerse myself with the fundamentals and avoid reliance.

This project will use the dataset from Figshare: click [here](https://figshare.com/articles/dataset/2022_Presidential_Elections_Data/19755469?file=35101492) to download the dataset.

## Setting Up the Environment

```{r Setting Up the Environment}
install.packages('tidyverse')
install.packages('readr')
install.packages('janitor')
install.packages('ggplot2')
install.packages('plotly')

library(tidyverse)
library(readr)
library(janitor)
library(ggplot2)
library(plotly)
```

## Reading the Dataset

Importing the dataset to the dataframe. In this case, the dataset is named as presidential_election.csv, feel free to change it to your preference!

```{r Data Frame Iniation}
pres_df <- data.frame(read_csv("presidential_election.csv"))

```


## Studying the Dataset

Here, I studied the structure of the dataset and the connection of each column to one another.

```{r Dataset Structure}
# head(pres_df)
# str(pres_df)
colnames(pres_df)
```

## Analysis 

This section will now take a closer look on the results of the 2022 election. Here, I added a dataframe that shows the candidates' names along with their campaign colors just for reference.

```{r Candidate Names and Campaign Colors}
candidates_df <- data.frame(
  id = 0:9,
  colname = c(
    "pres_1_abella",
    "pres_2_de_guzman",
    "pres_3_domagoso",
    "pres_4_gonzales",
    "pres_5_lacson",
    "pres_6_mangondato",
    "pres_7_marcos",
    "pres_8_montemayor",
    "pres_9_pacquiao",
    "pres_10_robredo"
  ),
  candidate_name = c(
    "Ernesto Abella",
    "Leody de Guzman",
    "Isko Moreno Domagoso",
    "Norberto Gonzales",
    "Panfilo Lacson",
    "Faisal Mangondato",
    "Ferdinand Marcos Jr.",
    "Jose Montemayor Jr.",
    "Manny Pacquiao",
    "Leni Robredo"
  ),
  campaign_color = c(
    "#808080",
    "#FFA500",
    "#0074D9",
    "#228B22",
    "#DDDDDD",
    "#8B4513",
    "#D50000",
    "#90EE90",
    "#FFD700",
    "#FF69B4"
  ),
  stringsAsFactors = FALSE
)

colnames(candidates_df)
```
Next, I want a quick summary of the total votes per candidate during the election.Since I have a `pres_df` containing their candidate names and notice that colname serves as the reference for both dataframe. Then, I added a column in the `candidates_df` to count all their corresponding votes throughout the country. 

In this chunk, each row will be checked out based on the according `colname` from `candidates_df`. This statement translates into "sum all votes from this column."

```{r Quick Preview}
candidates_df <- candidates_df %>%
  rowwise() %>%
  mutate(total_votes = sum(pres_df[[colname]])) %>%
  ungroup() 

candidates_df %>%
  select(candidate_name, total_votes) %>%
  arrange(desc(total_votes))
```

Now, let's visualize that.

```{r Plotting the preview}
ggplot(candidates_df, aes(x = reorder(candidate_name, -total_votes),
                                  y = total_votes,
                                  fill = I(campaign_color))) + 
  geom_bar(stat = "identity") +
  scale_fill_identity() + 
  scale_y_continuous(labels = comma) +
  labs(
    title = "2022 Philippine Presidential Candidate",
    x = "Candidate",
    y = "Total Votes"
  ) +
  theme (
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5, face = "bold")
  )
```

