---
title: "2022 Presidential Elections Data"
output: html_notebook
---

This is a personal project that aims to analyze the Philippine 2022 Presidential Election using the programming language, R. This project is based on Project Raven's Tableau [project](https://www.facebook.com/61562294530876/posts/pfbid02bwLWd45JfRSiCbmaFYBJRfM4WbYD6diK8rBAHRzoDjDwoZF8DWW8HZJ6gKJRhp9tl/). 

My personal aim in achieving this project is to avoid using any AI for assistance to fully immerse myself with the fundamentals and avoid reliance.

This project will use the dataset from Figshare: click [here](https://figshare.com/articles/dataset/2022_Presidential_Elections_Data/19755469?file=35101492) to download the dataset.

## Setting Up the Environment

```{r environment_set_up}
install.packages('tidyverse')
install.packages('readr')
install.packages('janitor')
install.packages('ggplot2')

library(tidyverse)
library(readr)
library(janitor)
library(ggplot2)

```
## Reading the Dataset

Importing the dataset to the dataframe. In this case, the dataset is named as presidential_election.csv, feel free to change it to your preference!

```{r data_frame_initiation}
pres_df <- data.frame(read_csv("presidential_election.csv"))

```


## Studying the Dataset

Here, I studied the structure of the dataset and the connection of each column to one another.

```{r}
head(pres_df)
str(pres_df)
colnames(pres_df)
```


